xolotl: An Intuitive and Approachable Neuron \& Network Simulator

### 2018-08-15 paper/paper.tex
working on supp figure comparison to ode23t
minor changes to paper; still working on benchmarks supp

	Conductance-based models of neurons are used extensively in computational neuroscience, and working with these models continues to be a challenge due to their high dimensionality and large number of parameters. Here, we present a neuron and network simulator built on a novel automatic type system that binds object-oriented code written in C++ to objects in MATLAB. Our approach builds on the tradition of uniting the speed and power of languages like C++ with the ease-of-use and feature-set of scientific programming languages like MATLAB. Our software allows for the creation and manipulation of hierarchical models with components that are named and searchable, permitting intuitive high-level programmatic control over all parameters. Our simulator's architecture allows for the live manipulation of any parameter in any model, and for visualizing the effects of changing that parameter immediately. Our simulator is fully featured with hundreds of ion channel models from the electrophysiological literature, and can be extended to include arbitrary conductances, synapses, and mechanisms. Several core features like bookmarking of parameters and automatic hashing of source code promote reproducible and auditable research. Finally, our simulator is written in a modular fashion and has been released under a permissive free software license, enabling seamless integration into the workflow of other researchers.

@@

Two major approaches have dominated the design of neuron simulators. One approach is to write the simulator in a fast compiled language like C and allow for the construction and simulation of neuron models using object-oriented paradigms. This approach has been implemented in NEURON (\cite{hinesNEURONSimulationEnvironment1997}). Simulators designed in this way tend to perform fast computations with little overhead, but suffer from a steep learning curve. Wrapping these simulators in a more approachable language like "Python"  or using graphical user interfaces (GUIs) mitigates these drawbacks (\cite{hinesNEURONPython2009}) at the cost of obfuscating the underlying algorithms and parameters (\cite{bretteSimulationNetworksSpiking2007, hinesNEURONPython2009}). In contrast, simulators designed from the ground up in popular scientific computing languages can be easier to use and benefit from interoperability with other commonly-used tools. Simulators like DynaSim (\cite{sherfeyDynaSimMATLABToolbox2018}), "ANNarchy" (\cite{vitayANNarchyCodeGeneration2015}), "BRIAN" (\cite{stimbergBrianSecondComing2013}), "morphforge" (\cite{hullMorphforgeToolboxSimulating2014}), and "PyNN" (\cite{davisonPyNNCommonInterface2009}) allow the user to specify models with strings of equations or components that are constructed using a special syntax, that can then be translated into a faster implementation language such as "C" or C++ (\cite{stimbergEquationorientedSpecificationNeural2014}). This approach permits considerable flexibility for simulating systems of differential equations. Since models need to be translated between the two languages, the hierarchical nature of neuron models is not naturally encapsulated by these tools, and the syntax can be verbose. Neither approach facilitates the creation of tools that simultaneously maintain efficiency, ease-of-use, and clarity.

To overcome these design limitations, we have developed a novel automatic type system, that we call "cpplab", which binds MATLAB code to classes specified in C++ header files. This architecture automatically creates objects in MATLAB that represent the underlying object-oriented C++ code, allowing the symbolic manipulation of C++ objects in the MATLAB interface. In this paper, we introduce xolotl, an implementation of the "cpplab" system specialized in integrating conductance-based neuron and network models. Models can be easily constructed from network building blocks in a few lines of MATLAB code using a hierarchical and intuitive syntax. Since models in the MATLAB workspace are automatically linked to models in the C++ implementation, configuring these objects in MATLAB transparently configures the underlying C++ objects. xolotl comes packaged with hundreds of components that can be used to assemble cells and networks;  has built-in visualization functions to inspect voltage time traces and activation functions; and a GUI for real-time manipulation of model parameters. The ease-of-use of the tools built into our software lends them to pedagogical applications and the rapid prototyping of models. Our software aims to simplify the investigation of the dynamics of conductance-based network and neuron models, facilitate collaborative modeling, and is intended to complement other tools being developed in the computational neuroscience community.

@@

The axon of the giant squid contains contains a fast inactivating sodium conductance ("NaV"), a slower non-inactivating potassium conductance ("Kd"), and a passive leak current ("Leak"). Seminal work by Hodgkin and Huxley showed that depolarizations of the membrane could lead to an activation of the voltage-sensitive "NaV" channels, which led to a run away depolarization that was terminated by the inactivation of "NaV" channels and the activation of "Kd" channels that repolarized the membrane (\cite{hodgkinComponentsMembraneConductance1952, hodgkinMeasurementCurrentvoltageRelations1952}). As one of the simplest models of excitable neural membranes, the Hodgkin-Huxley model often serves as a the first model introduced in pedagogical literature (\cite{dayanTheoreticalNeuroscience2001, sterrattPrinciplesComputationalModelling2011, trappenbergFundamentalsComputationalNeuroscience2010}).

In this example, we demonstrate how to simulate the spiking activity of a Hodgkin-Huxley-like model, and how the tools built into our simulator make it easy to set up and integrate the model and gain insight into the underlying biophysical mechanisms. This simple model consists of a single electrical compartment with three types of conductances (Figure \ref{fig:figurehh}A). This hierarchical organization of the neuron is mirrored in the structure of the model in our simulator: an object of type "compartment" is used to represent the cell body, and three objects of type "conductance" are used to model the three populations of ion channels (Figure \ref{fig:figurehh}B). These models of ion channels were obtained from \cite{liuModelNeuronActivityDependent1998} based on electrophysiological recordings from the lobster stomatogastric ganglion (\cite{turrigianoSelectiveRegulationCurrent1995}), and are part of the simulator. The code to set up this model is terse, idiomatic, relies on no special markup, and preserves the heirarchical nature of the model (Figure \ref{fig:figurehh}C).

@@

Neurons communicate and interact using synapses, electrochemical junctions between cells (\cite{gjorgjievaComputationalImplicationsBiophysical2016, huaNeuralActivityDynamics2004}). In chemical synapses, the presynaptic neuron releases packets of neurotransmitter across the synaptic cleft, which activate receptors on the postsynaptic neuron. In electrical synapses, no chemical intermediary is involved. Ions flux directly from the presynaptic neuron to the postsynaptic one, setting up a current in the post-synaptic cell. New patterns of activity can emerge from the characteristics of the connecting synapses (\cite{liComodulationSynapsesVoltagegated2018, nadimSynapticDepressionCreates1999, gutierrezRectifyingElectricalSynapses2013, gutierrezMultipleMechanismsSwitch2013}).

@@

Our software is dfesigned to streamline this process and allows for any parameter in any model to be manipulated using graphical sliders, with immediate, real-time feedback of its behavior. Any model in our simulator can be manipulated using

@@

We set out to design a neuron and network simulator that could be useful in the classroom setting, especially for students of computational science, while also being powerful, fast, and extensible enough to be used for research. By using a novel architecture that permits the symbolic manipulation of C++ objects in a intuitive MATLAB interface, we demonstrated some of the features of our simulator using simulations of single-compartment models of Hogkin-Huxley like neurons (Fig. \ref{fig:figurehh}); voltage clamp experiments to recover activation functions of single channel types (Fig. \ref{fig:figureclamp}); a neuron model where intracellular mechanisms can control the dynamics of Calcium and can regulate the maximal conductance of ion channel types in an activity-dependent manner (Fig. (\ref{fig:figuremechanism}); and a network of neurons with multiple synapse types (Fig. \ref{fig:figurenetwork}). We also illustrated how built-in features of our simulators make it easy to bookmark and jump between model configurations (Fig. \ref{fig:figuremechanism}, purple), and how parameters of the model can be changed using sliders and their effect can be viewed in real time (Fig. \ref{fig:figuremanipulate}).

>>>>

	Conductance-based models of neurons are used extensively in computational neuroscience, and working with these models continues to be a challenge due to their high dimensionality and large number of parameters. Here, we present a neuron and network simulator built on a novel automatic type system that binds object-oriented code written in C++ to objects in MATLAB. Our approach builds on the tradition of uniting the speed of languages like C++ with the ease-of-use and feature-set of scientific programming languages like MATLAB. Our software allows for the creation and manipulation of hierarchical models with components that are named and searchable, permitting intuitive high-level programmatic control over all parameters. Our simulator's architecture allows for the interactive manipulation of any parameter in any model, and for visualizing the effects of changing that parameter immediately. Our simulator is fully featured with hundreds of ion channel models from the electrophysiological literature, and can be extended to include arbitrary conductances, synapses, and mechanisms. Several core features like bookmarking of parameters and automatic hashing of source code facilitate reproducible and auditable research. Finally, our simulator is written in a modular fashion, includes detailed tutorials and worked examples, and is freely available at https://github.com/sg-s/xolotl, enabling seamless integration into the workflow of other researchers.

@@

Two major approaches have dominated the design of neuron simulators. One approach is to write the simulator in a fast compiled language like C and allow for the construction and simulation of neuron models using object-oriented paradigms. This approach has been implemented in NEURON (\cite{hinesNEURONSimulationEnvironment1997}). Simulators designed in this way tend to perform fast computations with little overhead, but suffer from a steep learning curve. Wrapping these simulators in a more approachable language like Python or using graphical user interfaces (GUIs) mitigates these drawbacks (\cite{hinesNEURONPython2009, gratiy2018bionet}) at the cost of obfuscating the underlying algorithms and parameters (\cite{bretteSimulationNetworksSpiking2007, hinesNEURONPython2009}). In contrast, simulators designed from the ground up in popular scientific computing languages can be easier to use and benefit from interoperability with other commonly-used tools. Simulators like DynaSim (\cite{sherfeyDynaSimMATLABToolbox2018}), ANNarchy (\cite{vitayANNarchyCodeGeneration2015}), BRIAN (\cite{stimbergBrianSecondComing2013}), morphforge (\cite{hullMorphforgeToolboxSimulating2014}), and PyNN (\cite{davisonPyNNCommonInterface2009}) allow the user to specify models with strings of equations or components that are constructed using a special syntax, that can then be translated into a faster implementation language such as C or C++ (\cite{stimbergEquationorientedSpecificationNeural2014}). This approach permits considerable flexibility for simulating systems of differential equations. Since models need to be translated between the two languages, the hierarchical nature of neuron models is not naturally encapsulated by these tools, and the syntax can be verbose. Neither approach facilitates the creation of tools that simultaneously maintain efficiency, ease-of-use, and clarity.

To overcome these design limitations, we have developed a novel automatic type system, that we call cpplab, which binds MATLAB code to classes specified in C++ header files. This architecture automatically creates objects in MATLAB that represent the underlying object-oriented C++ code, allowing the symbolic manipulation of C++ objects in the MATLAB interface. In this paper, we introduce xolotl, an implementation of the cpplab system specialized in integrating conductance-based neuron and network models. Models can be easily constructed from network building blocks in a few lines of MATLAB code using a hierarchical and intuitive syntax. Since models in the MATLAB workspace are automatically linked to models in the C++ implementation, configuring these objects in MATLAB transparently configures the underlying C++ objects. xolotl comes packaged with hundreds of components that can be used to assemble cells and networks;  has built-in visualization functions to inspect voltage time traces and activation functions; and a GUI for real-time manipulation of model parameters. The ease-of-use of the tools built into our software lends them to pedagogical applications and the rapid prototyping of models. Our software aims to simplify the investigation of the dynamics of conductance-based network and neuron models, facilitate collaborative modeling, and is intended to complement other tools being developed in the computational neuroscience community.

@@

The axon of the giant squid contains a fast inactivating sodium conductance ("NaV"), a slower non-inactivating potassium conductance ("Kd"), and a passive leak current ("Leak"). Seminal work by Hodgkin and Huxley showed that depolarizations of the membrane could lead to an activation of the voltage-sensitive "NaV" channels, which led to a run away depolarization that was terminated by the inactivation of "NaV" channels and the activation of "Kd" channels that repolarized the membrane (\cite{hodgkinComponentsMembraneConductance1952, hodgkinMeasurementCurrentvoltageRelations1952}). As one of the simplest models of excitable neural membranes, the Hodgkin-Huxley model often serves as a the first model introduced in pedagogical literature (\cite{dayanTheoreticalNeuroscience2001, sterrattPrinciplesComputationalModelling2011, trappenbergFundamentalsComputationalNeuroscience2010}).

In this example, we demonstrate how to simulate the spiking activity of a Hodgkin-Huxley-like model, and how the tools built into our simulator make it easy to set up and integrate the model and gain insight into the underlying biophysical mechanisms. This simple model consists of a single electrical compartment with three types of conductances (Figure \ref{fig:figurehh}A). This hierarchical organization of the neuron is mirrored in the structure of the model in our simulator: an object of type "compartment" is used to represent the cell body, and three objects of type "conductance" are used to model the three populations of ion channels (Figure \ref{fig:figurehh}B). These models of ion channels were obtained from \cite{liuModelNeuronActivityDependent1998} based on electrophysiological recordings from the lobster stomatogastric ganglion (\cite{turrigianoSelectiveRegulationCurrent1995}), and are part of the simulator. The code to set up this model is terse, idiomatic, relies on no special markup, and preserves the hierarchical nature of the model (Figure \ref{fig:figurehh}C).

@@

Neurons communicate and interact using synapses, electrochemical junctions between cells (\cite{gjorgjievaComputationalImplicationsBiophysical2016, huaNeuralActivityDynamics2004}). In chemical synapses, the presynaptic neuron releases packets of neurotransmitter across the synaptic cleft, which activate receptors on the postsynaptic neuron. In electrical synapses, no chemical intermediary is involved. New patterns of activity can emerge from the characteristics of the connecting synapses (\cite{liComodulationSynapsesVoltagegated2018, nadimSynapticDepressionCreates1999, gutierrezRectifyingElectricalSynapses2013, gutierrezMultipleMechanismsSwitch2013}).

@@

Our software is designed to streamline this process and allows for any parameter in any model to be manipulated using graphical sliders, with immediate, real-time feedback of its behavior. Any model in our simulator can be manipulated using

@@

We set out to design a neuron and network simulator that could be useful in the classroom setting, especially for students of computational science, while also being powerful, fast, and extensible enough to be used for research. By using a novel architecture that permits the symbolic manipulation of C++ objects in a intuitive MATLAB interface, we demonstrated some of the features of our simulator using simulations of single-compartment models of Hodgkin-Huxley like neurons (Fig. \ref{fig:figurehh}); voltage clamp experiments to recover activation functions of single channel types (Fig. \ref{fig:figureclamp}); a neuron model where intracellular mechanisms can control the dynamics of Calcium and can regulate the maximal conductance of ion channel types in an activity-dependent manner (Fig. (\ref{fig:figuremechanism}); and a network of neurons with multiple synapse types (Fig. \ref{fig:figurenetwork}). We also illustrated how built-in features of our simulators make it easy to bookmark and jump between model configurations (Fig. \ref{fig:figuremechanism}, purple), and how parameters of the model can be changed using sliders and their effect can be viewed in real time (Fig. \ref{fig:figuremanipulate}).
